---
title: "STAT 240 Discussion 10"
output: html_document
---

## Group 319C 

## LILY FRANKS, **HAOXUAN LU**, SHRUTHI SRINIVASAN, MICHAEL HANDLER

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## Questions


## Data

The following code reads in the Madison weather data and calculates the average July temperature for each year.
Treat the years from 1869 to 1898 as *early Madison* and from 1991 to 2020 as *recent Madison*.

```{r read-data, echo=FALSE}
## Read and transform the Madison weather data
mw <- read_csv("../../data/madison-weather-2020-clean-2.csv")

mw_july <- mw %>%
  filter(month=="Jul") %>%
  filter(year < 1899 | year > 1990) %>%
  group_by(year) %>%
  summarize(tavg = mean(tavg)) %>%
  mutate(era = case_when(
    year < 1900 ~ "early",
    year > 1989 ~ "recent"))
#view(mw_july)
```

### 1 

> Calculate the mean and standard deviation of the average July temperature separately for early Madison and recent Madison.

```{r problem-1}
q1 = mw_july%>%
  group_by(era)%>%
  summarize(mean = mean(tavg), sd = sd(tavg))
q1
```



### 2

> Make a plot with separate density curves for the early and recent July average temperatures.

```{r problem-2}
mw_july%>%
  group_by(era)%>%
  ggplot(aes(x = tavg, color = era)) +
  geom_density()
```




### 3

> Treat the 30 years in each time period as separate samples. Use the formulas in the instructions HTML to find 95% confidence intervals for the average July temperature in each time period. Compare results with the output of the `t.test()` function in the chunk below.

```{r problem-3a}

x1 <- mw_july %>%
  filter(era=="early") %>%
  pull(tavg)

t.test(x1)


x2 <- mw_july %>%
  filter(era=="recent") %>%
  pull(tavg)

t.test(x2)
```

```{r problem-3}
#early era CI
# sd
s1 = q1%>%
  filter(era=="early")%>%
  pull(sd)

#xbar
xbar1 = q1%>%
  filter(era=="early")%>%
  pull(mean)

# t*
x1t = qt(.025, 29)

early_CI <- tibble(early_UCL = xbar1 + (x1t * (s1 / sqrt(30))),
                   early_LCL = xbar1 - (x1t * (s1 / sqrt(30)))
                   )
early_CI

#recent era CI

# sd
s2 = q1%>%
  filter(era=="recent")%>%
  pull(sd)

#xbar
xbar2 = q1%>%
  filter(era=="recent")%>%
  pull(mean)

recent_CI <- tibble(recent_UCL = xbar2 + (x1t * (s2 / sqrt(30))),
                    recent_LCL = xbar2 - (x1t * (s2 / sqrt(30)))
                    )
recent_CI



```
The results of the confidence intervals presented in the one sample t-test output of the t.test() function is reflected in the confidence intervals we constructed. 


### 4

> Draw a graph of a t distribution with 29 degrees of freedom and with the middle 95% of the distribution shaded in.

```{r problem-4}

gt(df = 29)+
 geom_t_fill(df= 29, a=qt(.975, 29), b=qt(.025, 29), fill = "skyblue") +
  labs(title = "t-distribution", 
       subtitle = "DF = 29°")

```



### 5

> Find a 95% confidence interval for the change (early minus recent) of average July temperatures in Madison.
Calculate the difference in sample means and calculate the standard error for the difference.
Use the `t.test()` function result to find the estimated degrees of freedom.
Find the critical $t^*$ value for a 95% confidence interval for this number of degrees of freedom.
Verify that the hand calculation for the confidence interval matches the `t.test()` result.

```{r problem-5}
t.test(x1,x2)

x5_a <- mean(x1) - mean(x2) + qt(0.025,54.733) * sqrt(sd(x1)^2/30 + sd(x2)^2/30)
x5_a

x5_b <- mean(x1) - mean(x2) + qt(0.975,54.733) * sqrt(sd(x1)^2/30 + sd(x2)^2/30)
x5_b
```



### 6

> Test the hypothesis that the mean July temperature in Madison in the early era is equal to that in the recent era versus that alternative that they are different.
Calculate the test statistic $t$ from the equations
and verify it matches the result from `t.test()`. Write a conclusion in context to interpret the result of this hypothesis test.

```{r problem-6}
t.test(x1,x2)

t_value <- (mean(x1) - mean(x2)) / sqrt(sd(x1)^2/30 + sd(x2)^2/30)
t_value

p <- pt((mean(x1) - mean(x2)) / sqrt(sd(x1)^2/30 + sd(x2)^2/30), 54.733, lower.tail = FALSE)*2
p
```

- Because p_value=0.2889>0.05, we failed to reject the null hypothesis


### 7

> Make a plot of the t distribution for the previous problem. Add a vertical line at the test statistic and shade in the two tail areas that correspond to the p-value.

```{r problem-7}
gt(54.733)  +
  geom_t_fill(54.733, a = -4,b = -t_value) +
  geom_t_fill(54.733, a = t_value, b = 4) +
  geom_vline(xintercept = t_value)
```



