---
title: "STAT 240 Discussion 8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

## Group 4 DIS 314

## Members Present

- Wai Zin Linn, Will Hu
- **ZHAN YU**

## Members Absent

- Martin Lozano

## Questions

### 1

(a) A fair coin is tossed repeatedly until the tenth head is tossed. $X_1$ is the number of tails tossed prior to the tenth head.


-This does not fit in binomial model, because the number of trials isn't decided before the experiment. 


(b) Four buckets each contain a total of five tokens each, some red and some blue. The number of red tokens in the buckets are 1, 2, 3, and 4 with blue tokens making up the remainder. One token is drawn at random from each bucket. $X_2$ is the total number of red tokens drawn.

-This does not fit in binomial model, because the trials don't have the same probability of success.


(c) A random sample of students in a fitness study. $X_3$ is the mean systolic blood pressure of the sample.

-This does not fit in binomial model, because the blood pressure is a continuous variable, and the result is not binomial.

### 2

Create a data frame with columns `n`, `p`, `x`, `mu`, and `prob` where:

- `n` varies from 2 to 100 by twos (so each `n` value is even:  2, 4, 6, 8, ..., 100);
- `p` equals 0.5;
- `x` is `n`/2;
- `mu` is the mean of the Binomial distribution with parameters `n` and `p`;
- `prob` is $P(X = x)$, where X~Binomial(`n`, `p`)

Show the first few rows using `head()`, then respond to the question below.


```{r}
q2 <- tibble(
  n = seq(2, 100, 2),
  p = 0.5,
  x = n / 2,
  mu = n * p,
  prob = dbinom(x, n, p)
)

head(q2)
```
What is the relationship between `x` and `mu`?

-They are the same


### 3

The central limit theorem implies that the binomial distribution converges to a normal distribution as $n$ increases.  This problem will examine one aspect of the convergence, namely the right tail probability of being more than two standard deviations above the mean, $P(X > \mu + 2\sigma)$.

**(a)** What is the probability that a normal random variable with mean $\mu$ and standard deviation $\sigma$ exceeds $\mu + 2 \sigma$? Display answer rounded to four decimal places.

```{r}
round(1 - pnorm(2, 0, 1), 4)
```

Response: 0.0228
**(b)** 

Create a data frame with columns `n`, `p`, `mu`, `sigma`, `x`, and `prob` where:  
- `n` varies from 1 to 1000 by ones (1, 2, 3, ..., 1000);  
- `p` equals 0.5;  
- `mu` is the mean of the Binomial distribution with parameters `n` and `p`;
- `sigma` is the standard deviation of the Binomial distribution with parameters `n` and `p`;
- `x` equals `mu` + 2*`sigma`;  
- `prob` is $P(X > x)$, where X~Binomial(`n`, `p`)

Display the first few rows of the data frame with `head()`.
```{r}
q3_0.5 <- tibble(
  n = seq(1, 1000, 1),
  p = 0.5,
  mu = n * p,
  sigma = sqrt(n * p * (1-p)),
  x = mu + 2*sigma,
  prob = 1 - pbinom(x, n, p)
)
head(q3_0.5)
```

**(c)** Plot the right tail probabilities versus $n$, and carryout the following modifications:

- Add a red, dashed, horizontal line at the value you found in part **(a)**.  
- Add a smooth curve which lessens the visual impact of the oscillations due to the discreteness of the binomial distribution.
```{r}
ggplot(q3_0.5, mapping = aes(x = n, y = prob)) + 
  geom_line() +
  geom_hline(aes(yintercept = round(1 - pnorm(2, 0, 1), 4)), linetype="dashed", color = "red") +
  geom_smooth()
```

**(d)** Repeat parts **(b)** and **(c)** if $p = 0.005$.
```{r}
q3_0.005 <- tibble(
  n = seq(1, 1000, 1),
  p = 0.005,
  mu = n * p,
  sigma = sqrt(n * p * (1-p)),
  x = mu + 2*sigma,
  prob = 1 - pbinom(x, n, p)
)
head(q3_0.005)
```

```{r}
ggplot(q3_0.005, mapping = aes(x = n, y = prob)) + 
  geom_line() +
  geom_hline(aes(yintercept = round(1 - pnorm(2, 0, 1), 4)), linetype="dashed", color = "red") +
  geom_smooth()
```

**(e)**

Ignoring the oscillations, how do the patterns of the two smooth curves in the graphs differ from one another, especially when $n$ is large?

-When $n$ is larger, the smooth curves close to the value we found in part a more quickly as the n increase. 

### 4

Draw a graph of the binomial distribution with $(n=500, p = 0.5)$ and a graph for the binomial distribution with $(n=500, p = 0.005)$; you may use the `gbinom` function in the *ggprob.R* script.
Scale these graphics so that the x axis is restricted to where the probabilities are relatively large; set `scale=TRUE` in the `gbinom` function to accomplish this.  

Overlay each plot with a red normal density with a mean and standard deviation that matches the mean and standard deviation of the corresponding binomial distribution.  Consider the `geom_norm_density` function in the *ggprob.R* script.

Compare the skewness of the distributions and comment on how this might help explain the differences in the right tail probabilities from Problem 3.
-Answer: Due to its bell shape, Binomial(500, 0.5) has a normal distribution, whereas Binomial(500, 0.005) has very few trials and is therefore positively skewed. 

```{r}
gbinom(500, 0.5, a = 200, b = 300) + 
  geom_norm_density(500*0.5, sqrt(500 * 0.5 * (1-0.5)), scale = TRUE, color = "red") + 
  xlab("n")

gbinom(500, 0.005, a = -5, b = 9) + 
  geom_norm_density(500*0.005, sqrt(500 * 0.005 * (1-0.005)), scale = TRUE, color = "red") + 
  xlab("n")
```





