---
author: "Zhan Yu"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
library(lubridate)
```

## Assignment 2


- Execute the code in the chunk below to read in the data.
    - If there is an error, check to see if there is a `data` directory two directories up that contains the data file.
- Post questions on our Piazza site, visit the Learning Center, or attend office hours if you have questions.
    
```{r read-data, echo = FALSE}
## The echo = FALSE argument to this chunk
##   means that the code will not appear in the output
##   but the code will run so that the data is in the session

## Read Lake Monona data
## Remove rows with missing data
monona <- read_csv("../../data/lake-monona-clean-2020.csv") %>%
  drop_na()
```

### Problems

### 1

The following code makes a histogram of the `days` variable in the Lake Monona data set.

```{r problem1}
ggplot(monona, aes(x=days)) +
  geom_histogram(boundary=0,binwidth=10,color="black",fill="white")
```

Approximately how many winters had between 40 to 70 days where Lake Monona was at least 50% covered with ice?

### Response

11 days


### 2

Modify the code below so that:

- one of the bin boundaries is at 70 days
- the width of each bin is 5 days
- the fill color is "cyan"
- the color outlining the bars is "red"
- the x label says "Days Closed with Ice"
- the y label says "Total"
- there is a title with words of your choosing that describe the figure

```{r problem2}
ggplot(monona, aes(x = days)) +
  geom_histogram(boundary=70,binwidth=5,color="red",fill="cyan")+
  ggtitle("Ice Influence")+
  xlab("Days Closed with Ice") + 
  ylab("Total")
```


### 3

Code in the next chunk makes a scatter plot that shows how the variable `days` changes with time (using `year1`).

```{r problem3}
ggplot(monona, aes(x = year1, y = days)) +
  geom_point() +
  geom_smooth(se=FALSE)
```

- What does the line of code `geom_smooth(se=FALSE)` do?  (Explain what it does on the graphic; you don't need to explain details of the method.)
- Describe the pattern of the curve.
- How long was Lake Monona closed with ice in a typical year near 1875 (i.e., what is the approximate value of the smooth curve around 1875)?
- How long was Lake Monona closed with ice in a typical year near 2000 (i.e., what is the approximate value of the smooth curve around 2000)?

### Response

- Add a smooth trend line.
- Increasing with years, the amount of days closed with ice in the year is continuously fluctuating but showing a downward trend.
- 118
- 90


### 4

Modify the code in the following chunk so that:

- There is a box plot displaying the distribution of the days frozen by ice  
- The box plot fill color is "yellow"
- The color of the edges of the box plot is "magenta"
- There is a more descriptive y-axis label
- There is an informative plot title

```{r problem4}
ggplot(monona, aes(y=days)) +
  geom_boxplot(color="magenta",fill="yellow") +
  ylab("Total days frozen")+
  ggtitle("Lake Monona Freeze Durations, 1851-2020")
```

- What is the approximate median number of days Lake Monona has been closed with ice?  

### Response

105

f