---
title: "STAT 240 Discussion 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
source("../../scripts/viridis.R")
```

## Group 317C 

## Members Present

- LILY FRANKS, SHRUTHI SRINIVASAN, MICHAEL HANDLER
- **Haoxuan Lu**

## Members Absent

- ADD NAMES OF ANY ABSENT MEMBERS


## Questions


### 1

Complete a function that takes as input a sample size `n`, a mean `mean`, and a standard deviation `sd`, and does the following:   
  - generates a random sample of size `n` from a normal distribution with mean `mean` and standard deviation `sd`  
  - calculates the sample mean  
  - calculates the difference between the sample mean and the mean of the distribution  
  - returns the calculated difference

```{r problem-1}
## Here is code to do these steps without a function
##   for n=100, mean=50, sd=10
mu <- 50
sigma <- 10
## What does the next line of code do?  list 100 random samples of normal distribution values with mean 50 and standard deviation 10 and store the result in the variable x
x <- rnorm(n=100, mean=mu, sd=sigma)

mean(x) - mu

## To repeat this code many times, it will be helpful to have a function.
## We will not  use this function a lot, so pick a generic name.
## The function has default values
## Replace the XXX with arguments to rnorm()
f_1 <- function(n=1,mean=0,sd=1){
  x <- rnorm(n, mean, sd)
  return ( mean(x) - mean )
}
```



### 2

Use a `for` loop to run the previous function 1000 times with `n=100`, `mean=50` and `sd=10`. What are the mean and standard deviation of these differences? Repeat for the same mean and sd, but with `n=10000`. Compare the mean and standard deviation of the differences for this larger sample size.  The code has been started for you below.

```{r problem-2}
## Here is a for-loop
diff_100 <- numeric(1000)
for ( i in 1:1000 ){
  diff_100[i] <- f_1(100,50,10)
}

mean(diff_100)
sd(diff_100)

## ADD CODE TO REPEAT FOR n=10000
diff_10000 <- numeric(1000)
for ( i in 1:1000 ){
  diff_10000[i] <- f_1(10000,50,10)
}

mean(diff_10000)
sd(diff_10000)
```

### Response

How do the mean and sd of the difference between the sample mean and the input mean change when $n$ increases from 100 to 10,000?

- the sd of the difference between the sample mean and the input mean become smaller and the mean of the difference between the sample mean and the input mean become closer to 0 when $n$ increases from 100 to 10,000


### 3

The function `pnorm()` calculates the total area to the **left** of the given value for a normal curve with a given mean and standard deviation. Below are a couple examples. 
The difference between these probabilities represents the area between 50 and 60 under the given normal curve.

```{r problem-3a}
p1 <- pnorm(q=50, mean=50, sd=10)
p2 <- pnorm(q=60, mean=50, sd=10)

p1
p2
p2 - p1
```

Write a function with arguments `a`, `b`, `mean`, and `sd`
that finds the area between `a` and `b` under a normal density with the given mean and sd.

```{r problem-3}
interval_pnorm <- function(a,b,mean,sd){
  ## complete the code
  pa <- pnorm(a, mean, sd)
  pb <- pnorm(b, mean, sd)
  return(pb - pa)
}

## Test out your function
interval_pnorm(50,60,mean=50,sd=10)
```


### 4

Evaluate the following normal interval areas using your function from question 3.

- Between $-2$ and 2 when `mean=0` and `sd=1`.
- Between 60 and 140 when `mean=100` and `sd=20`.
- Between 90 and 110 when `mean=100` and `sd=5`.

```{r problem-4}
interval_pnorm(-2, 2, 0, 1)
interval_pnorm(60, 140, 100, 20)
interval_pnorm(90, 110, 100, 5)
```

Explain why all of these values are the same.

- Because all the interval is from (mean - 2*sd) to (mean + 2*sd), since they are all normal distributions examples, the interval areas have the same portion.


### 5

The function `qnorm()` finds a quantile from a normal distribution.
The first argument `p` is a number strictly between 0 and 1.
The function returns the quantile `q` where the total area to the left of `q` under the normal curve is equal to `p`.
The function `qnorm()` is the inverse of the function `pnorm()`.

#### Examples

```{r problem-5a}
qnorm(0.5) # Median of a N(0,1) is 0
qnorm(0.84, mean=50, sd=10)
qnorm(0.975, mean=50,sd=10)
```

Write a function that takes as input a number `p` between 0 and 1, a mean `mean`, and a standard deviation `sd` and returns two numbers `a` and `b` (use `c(a,b)`) so that `a` and `b` are equidistant from the mean and the area between `a` and `b` is equal to `p`.

For example, if `p=0.6`, `mean=50` and `sd=5`, then the area **not** between `a` and `b` is $1 - 0.6 = 0.4$. The area to the left of `a` is half this, or 0.2. The area to the right of `b` is also 0.2, which means the area to the left of `b` is 0.8.
Therefore,

```{r problem-5b}
a <- qnorm(0.2,50,5)
b <- qnorm(0.8,50,5)
c(a,b)
```

Write the function.

```{r problem-5}
interval_qnorm <- function(p,mean=0,sd=1){
  ## XXX write this code
  a <- qnorm(0.5 - p/2, mean, sd)
  b <- qnorm(0.5 + p/2, mean, sd)
  return(c(a,b))
}

## Test out your function
interval_qnorm(.6,50, 5)
```


### 6

Evaluate the previous function to find the end points of the intervals that contain the middle 95% (`p=0.95`) for the standard normal curve and two other normal distributions (i.e., two normal distributions where you pick their means and standard deviations).

```{r problem-6}
interval_qnorm(0.95, 0, 1)
interval_qnorm(0.95, 10, 7)
interval_qnorm(0.95, 15, 9)
```


### 7

The *z-score* of a number $x$ relative to a distribution with mean $\mu$ and standard deviation $\sigma$ is
$$
z = \frac{x - \mu}{\sigma}
$$
The number `z` is the number of standard deviations (positive or negative) that the value $x$ differs from the mean $\mu$.

Evaluate the z-scores for the end points of the intervals you found in problem 6.
```{r problem-7}
z_score <- function(x, mean, sd){
  return((x - mean) / sd)
}

z_score(1.959964, 0, 1)
z_score(23.719748, 10, 7)
z_score(32.639676, 15, 9)
```

Explain the similarities.

### Response

- the z-scores for the end points of the intervals of the three cases above are all 1.959964. It is because the three cases are all normal distribution containing the middle 95%, so the z-score for the ending points are the same.
