---
title: "STAT 240 Discussion 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
```

## Group Name 3 (use the assigned group name)

## Members Present
WAI ZIN LINN
CAROLINE HABERLAND-ERVIN
SHANE SPELLMAN
**ZHAN YU**
- ADD THE NAMES OF EVERYONE PRESENT
- MAKE THE NAME **BOLD** OF THE RESPONSIBLE PERSON TO UPLOAD THE FILES FOR THE GROUP

## Members Absent

- ADD NAMES OF ANY ABSENT MEMBERS

## Exoplanet Data

We will be working with data on exoplanets; exoplanets are planets orbiting stars outside our Solar System.  The data were pulled from the [NASA Exoplanet Archive](https://exoplanetarchive.ipac.caltech.edu/index.html) on September 3, 2020.

```{r read-data, echo = FALSE}
## Read in the csv file
## There is one row per exoplanet 
## Select some variables that we will work with and rename them
## Remove very massive planet (only to improve plot visuals for this assignment)
## Drop missing values; the remaining exoplanets will have estimates of both mass and radius
planets <- read_csv("../../data/exoplanets-3sept2020.csv") %>%
  filter(default_flag == 1) %>%
  select(pl_name, discoverymethod, disc_year, sy_pnum, pl_rade, pl_bmasse) %>%
  rename(planet=pl_name, method=discoverymethod, year=disc_year, number=sy_pnum, radius=pl_rade, mass=pl_bmasse) %>%
  drop_na() 
```

The variables we will be working with are defined below.  If you want more details, please see the exoplanet archive [data column definitions](https://exoplanetarchive.ipac.caltech.edu/docs/API_PS_columns.html) (using the column names in the raw data).

`planet` = Planet Name  
`method` =  Method by which the planet was first identified  
`year` = Discovery Year (the year it was discovered)  
`number` = Number of Planets  
`radius` = Planet Radius (units: Earth Radius)  
`mass` = Approximate or Minimum Planet Mass (units: Earth Mass)  



## Questions

### Question 1

> Make a scatter plot with `radius` (the estimate of the planet radius) on the x-axis and `mass` (the estimate of the planet mass) on the vertical axis.  The `scale_y_log10()` and `scale_y_log10()` are added to change the axis values to a `log10()` scale.  

```{r question-1}
## COMPLETE CODE TO MAKE THIS PLOT
ggplot(planets) +
  scale_y_log10() +
  scale_x_log10() +
  xlab("Radius") +
  ylab("Mass (Earth Mass)") +
  ggtitle("Exoplanets Mass vs. Radius",
          subtitle="Using NASA Explanet Archive Data")+
  geom_point(aes(x=radius,y=mass))
```



### Question 2

> Make a scatter plot like the previous problem, but use the `color` aesthetic for `method` (discovery method).

```{r question-2}
## COMPLETE CODE TO MAKE THIS PLOT
ggplot(planets) +
  scale_y_log10() +
  scale_x_log10() +
  xlab("Radius") +
  ylab("Mass (Earth Mass)") +
  ggtitle("Exoplanets Mass vs. Radius",
          subtitle="Using NASA Explanet Archive Data")+
  geom_point(aes(x=radius,y=mass,color = method))
```



### Question 3

- Which discovery method was used to detect the most exoplanets in our `planet` data frame?  
- What general pattern do you notice between planet mass and radius?  
- What is the estimated radius (approximately) of the most massive planet included on the plot?


#### Response

- Transit
- There is a slight positive relationship between planet mass and      radius

- 14.0


### Question 4

> Another way to display differences by discovery method is to separate the plot into multiple panels (or facets) and plot different subsets of the data in each.
> We are going to look at planet mass versus discovery year.
> Add a comment to say what the code on the next line does where indicated.


```{r question-4}
ggplot(planets, aes(x=year,y=mass)) +
  geom_point() +
  ## Creates a line of best fit for the points in the graph
  # Response: 
  geom_smooth(method="lm") +
  ## Creates a different graph for each method
  # Response:  
  facet_wrap(~method) +
  xlab("Discovery Year") +
  ylab("Mass (Earth Mass)") +
  ggtitle("Exoplanets Mass vs. Discovery Year",
          subtitle="Using NASA Explanet Archive Data")
```




