---
title: "Draft"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warnings=FALSE)
library(tidyverse)
library(lubridate)
library(dplyr)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```
## Member Present
-**Zhan Yu**, Will Hu, WAI ZIN LINN

## Member Absent
-Martin Lozano

## Draft

## ***Introduction***
-We chose to focus on suicide rate because it will be an interesting topic to explore and this topic contains a vast amount of data. We decided to narrow our focus down to some particular factors that can potentially influence suicide rate. We are curious about the relationship between suicide rate and GDP per capita of each country, gender, age groups, so we decided to explore data about the suicide rates in each age group and gender in each country. In this project, we plan to answer the question: 
1.what is the relationship between different countries' GDP per capita and their suicide rates.
2.what is the relationship between different genders and their suicide rates.
3.what is the relationship between different age groups and their suicide rates.
-Our thesis statement is that we will calculate the sum of variables that relevant to GDP and suicides rate and use hypothesis test to test our result. The result is that the country with the higher GDP per capita will have higher suicide rate. Male will have higher suicide rate than female. Moreover, people age 75+ have the highest suicide rate.

## ***Background***
-This data set is pulled from four other datasets linked by time and space. These datasets come from United Nations Development Program, World Bank, World Health Organization, and Suicide in the Twenty-First Century. 

-The variables country, sex,and year mean some countries, genders and years(1985-2016) used in the dataset.The variable age group include these five groups: 15-24,25-34,35-54,55-74,75+. The variables count of suicides, population, suicide rate(number of suicides/100K population), country-year composite key, HDI for year, gdpforyear, gdppercapita, generation (based on age grouping average) are self-explantory.

-Dataset Link: https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016

-There are many factors like age, country, gdp, and gender can influence the suicide rate, so we are going to investigate the relationship between these factors and suicide rate.

-we do not have unusual factors which may affect interpretation of results.

-For the rest of report, we intend to analyze the dataset and give graphs to visualize the data.

## ***Analysis***

- The relationship between different countries' GDP per capita and their suicide rates.
- This result is calculated by summing of suicides rate in 100k people of each country, and summing the capita GDP of each country. Then we use scatter plot to show the relationship between countries' GDP per capita and the total suicide rate in 100k people of each country.
-Evidence: The project presented on https://rpubs.com/beancounter/430586 shows the same result as ours.
```{r include=FALSE}
data <- read_csv("../../data/suicide_rate.csv")
country <- data%>%
  group_by(country)%>%
  rename(suicides_rate = 'suicides/100k pop',capita_gdp = 'gdp_per_capita ($)')%>%
  summarise(total_suicide_rate=sum(suicides_rate),total_capita_gdp=sum(capita_gdp))
country
ggplot(data=country,aes(y=total_suicide_rate,x=total_capita_gdp))+
  geom_point()+
  geom_smooth(method=lm)+
  theme(axis.text.x = element_text(angle = 0, hjust = 1))+
  ggtitle("Relationship between Total Suicide Rate and each country's GDP per capita")+
  ylab("Total Suicide Rate")+
  xlab("Total GDP per capita")
  
  
```

-
-
-The relationship between different age groups and their suicide rates.
-The result is calculated by the sum of suicides number in 100k people of each age group divided by the number of rows of each group.
```{r include=FALSE}
age_group <- data%>%
  group_by(age)%>%
  summarise(rows=n(),
            suicide=sum(suicides_no)/rows,
            total=sum(population)/rows,
            avg_suicide_num=suicide/total *100)
age_group

ggplot(data=age_group,aes(x=reorder(age,desc(avg_suicide_num)),y=avg_suicide_num,fill=age))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))+
  ggtitle("Average Suicide Rates for Each Age Group")+
  xlab("Age Group")+
  ylab("Suicide Percentage (%)")
```

-
-

-what is the relationship between different genders and their suicide rates.
-The result is calculated by the sum of suicides number in 100k people of each gender divided by the number of rows of each gender.
```{r include=FALSE}
gender <- data%>%
  group_by(sex)%>%
  summarise(rows=n(),suicide=sum(suicides_no)/rows,
            total=sum(population)/rows,
            avg_suicide_num=suicide/total *100)
ggplot(data=gender,aes(x=reorder(sex,desc(avg_suicide_num)),y=avg_suicide_num,fill=sex))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 0, hjust = 1))+
  ggtitle("Suicide Percentage for Each Gender")+
  xlab("Gender")+
  ylab("Suicide Percentage (%)")
```
-Using hypothesis test on male percentage suicide rate.

```{r include=FALSE}
male_test <- data%>%
  group_by(sex)%>%
  filter("male"==sex)%>%
  summarise(rows=n(),
            suicide=round(sum(suicides_no)/rows,0),
            total=round(sum(population)/rows,0),
            p = suicide/total)
y <- 0:male_test$total
y_extreme <- y[dbinom(y,male_test$total,0.1) < dbinom(male_test$suicide,male_test$total,0.1)+tol]
p_value <- sum(dbinom(y_extreme,male_test$total,0.1))
p_value
gbinom(male_test$total,0.0002,scale=TRUE,size=0.5)+
  geom_vline(xintercept = male_test$total * 0.0002 + (male_test$total * 0.0002 - male_test$suicide), color = "red", linetype = "dashed")+
  geom_vline(xintercept = male_test$suicide,color="red",linetype="dashed",size=0.5)+
  theme_bw()

```

## ***Discussion***
For the relationship between suicide rate and every country's GDP per capita, the higher GDP per capita, the higher suicide rate the country has. This result is showed by a linear regression. For the suicide rates in different age group, as people become older, their suicide rate will also become higher, so the age group 75+ has the highest suicide rate. For different genders, male has higher suicide rate than female. In addition, we use hypothesis test on male suicide percentage. The result shows that the p_value is 1.78835e-05, which is less than 0.05, which is statistically significant. There is strong evidence that the null hypothesis is wrong, and the true value of p is closer to the observed ratio of p^≈373/1800818. p is not 0.5, we just observed typical data, which is more credible than p=0.5, we saw a very unlikely result. There is very strong evidence (p≈0.000017, two-sided binomial test) that in an experimental environment, the suicide rate of male is above 0.02%.

In the graph, comparing the relationship between age group and suicide rate, there seems to be some data missing for certain age groups of some years for some countries. For Albania's 2015 suicide total number, all of the age group is 0. Therefore, some of the data may be missing or there is no sucide taking place in that year at Albania. The data also seems to skip out on some years. For example in Albania, the years 1990, 1991 are not present.

For potential future directions, we have new questions: 1.What other factors may influence the suicide rate? (May need to look at different or larger datatsets). 2.For country with high suicide rates, what factor contributes that most? The comparison of different countries suicide rates can be calculated differently, as we took an approach which looks at the average suicide rate for each country, while we could also look at which country has the highest total suicides, instead of using the average. The same is applicable for the comparison with gender and GDP. The relationship between suicide rate and GDP could also be computed using GDP per year instead of GDP per capita, which we used.One important fact to collect to refine our understanding would be to see how these three organizations collected this data, to better understand if there was any inherent bias or any inherent flaws with the collection, to see if our data is skewed in any particular direction.

To sum up, country with the higher GDP per capita, people age 75+ and male have higher suicide rate. We also get some evidence from internet to prove our conclusion.
